---
title: "Multinomial CI Prediction for 2020 Election"
author: "Shanghao Zhong"
date: "2020-11-06"
output:
  pdf_document: default
  html_notebook: default
---

Load package
```{r}
library(jsonlite)
library(lubridate)
library(MultinomialCI)
library(dplyr)
```

You also need to load the data. You should have the `all.state.dataset`.
```{r}
load("MultinomialCIProject2020.RData")
all.state.dataset <- get('all.state.dataset')
```


## Predifine function

The script concerns about the probability of a vote cast for a specific candidate.
Assuming the votes follow a multinomial distribution, 
it uses the `MultinomialCI` package to calculate the confidence interval
of the probability for each candidate.

```{r}
#' Calculate county level multinomial c.i.
#'
#' @param df a data frame with vote counts. 
#'           row.name is county name, 
#'           and each column is the number of votes for each candidate.
#' @param alpha The significance level for the confidence intervals. 
#'              Must be a real number in the interval [0, 1]
#' @return a data frame of confidence intervals for each candidate
#' @export
CI.by.county <- function(df, alpha) {
  df <- filter(df, df[[1]] > 0)
  candidate.names <- names(df)
  county <- row.names(df)
  mat <- t(apply(df, 1, as.numeric))
  
  CI.per.county <- function(vec) {
    CI.per.county <- c(multinomialCI(pmax(vec, 0), alpha))
  }
  cols <- apply(expand.grid(candidate.names, c('low', 'high')), 1, paste, collapse='.')
  all.ci <- t(apply(mat, 1, CI.per.county))
  
  output <- data.frame(all.ci, row.names = county)
  colnames(output) <- cols
  
  return(output)
}
```

With the probability of a vote cast for a candidate, the final projection is done by
$np \pm z\cdot \sqrt{n\cdot p\cdot (1-p)}$, where $p$ is the probability of a candidate
winning a vote, $n$ is the expected remaining vote according to NYT.

```{r}
#' Calculate the lower and the upper bound of the remaining votes given the 
#' probability.
#'
#' @param remaining A scalar or vector containing the remaining votes in the 
#'                  county.
#' @param prob A scalar or vector containing the probability of votes going to 
#'             the candidate in the county.
#' @param alpha confidence level
#'
#' @return a length 2 vector, first is the lower bound and the second is the 
#'         upper.
#' @export
range_est <- function(remaining, prob, alpha=0.95) {
  lo <- remaining*prob + qnorm((1-alpha)/2)*sqrt(remaining*prob*(1-prob))
  hi <- remaining*prob - qnorm((1-alpha)/2)*sqrt(remaining*prob*(1-prob))
  return(c(sum(lo, na.rm = TRUE), sum(hi, na.rm = TRUE)))
}
```


```{r}
#' Helper function to convert UTC time string to EST
#'
#' @param timestring string representation of time, NYT's time in ISO format
#'
#' @return a POSIXct object in EST
#' @export
#'
#' @examples to_EST('2020-11-07T01:46:10Z')
to.EST <- function(timestring) {
  with_tz(parse_date_time(timestring, 'ymd HMS'), 'EST')
}
```


## Pull in data from NYT's api

Need to change the `current.state.name` --- state name used to pull data from NYT's api.

It should be spelled out and all lower case and whitespace should be replace by `-`.

e.g. 

- New York: `new-york`
- Pennsylvania: `pennsylvania`
- Distinct of Columbia: `district-of-columbia`

It will store all the pulled data in `all.state.dataset`. Only new data will be stored.

```{r}
# This script will load data for the following state.
current.state.name <- 'arizona'

nyt.api <- paste(
  'https://static01.nyt.com/elections-assets/2020/data/api/2020-11-03/race-page/', 
  current.state.name, 
  '/president.json',
  sep = '')

results <- fromJSON(nyt.api)
current.update.time <- max(results$data$races$counties[[1]]$last_updated)

# if the state haven't been track yet, create a slot for that.
if (!current.state.name %in% names(all.state.dataset)) {
  update.type <- "New data"
  all.state.dataset[[current.state.name]] <- list()
  all.state.dataset[[current.state.name]][[current.update.time]] <- results$data$races$counties[[1]]
  previous.update.time <- 'N/A'
} else { # if the state has been track, see if this one is new update
  previous.update.time <- max(last(all.state.dataset[[current.state.name]])
                              $last_updated)
  if (current.update.time == previous.update.time) {
    update.type <- "No update"
  } else {
    update.type <- "New update"
    all.state.dataset[[current.state.name]][[current.update.time]] <- results$data$races$counties[[1]]
  }
}

rm(results)

by.county <- last(all.state.dataset[[current.state.name]])
old.by.county <- nth(all.state.dataset[[current.state.name]], 
                     max(1, length(all.state.dataset[[current.state.name]])-1))


all.votes <- data.frame(by.county$results, 
                        row.names = by.county$name)
old.all.votes <- data.frame(old.by.county$results, 
                            row.names = old.by.county$name)

mail.votes <- data.frame(by.county$results_absentee, 
                         row.names = by.county$name)
old.mail.votes <- data.frame(old.by.county$results_absentee, 
                             row.names = old.by.county$name)

cat(paste(update.type, 'in', current.state.name),
    paste("Previous Update:", 
          to.EST(max(old.by.county$last_updated))),
    paste("Current Update:", 
          to.EST(max(by.county$last_updated))),
    paste("Previous margin:", 
          sum(old.all.votes$bidenj) - sum(old.all.votes$trumpd)),
    paste("Current margin:", 
          sum(all.votes$bidenj) - sum(all.votes$trumpd)),
    sep='\n')
```

See the number of snapshots saved
```{r}
sapply(all.state.dataset, length)
```

See the time at which snapshots were taken (in EST)
```{r}
to.EST(names(all.state.dataset[[current.state.name]]))
```


## Estimate the probability for each candidate

You can change `using` to update how you want to estimate the probability.
Unhide one of them and hide to other to use.

first option: 

- Use the difference between old data and new data
- Best for predicting the most recent trend
- Doesn't work if the difference between old data and new data is small or non-representative

second option: 

- Use the mail.votes to predict old data and new data 
- Work the best if the mail votes is homogeneous throughout different time
- Doesn't work But the demographics within mail data can change over time

When `using` is incomplete, the probability of each candidate will based on `all.votes`


```{r}
# first option
using <- data.frame(
  data.matrix(last(all.state.dataset[[current.state.name]])$results) 
  - data.matrix(nth(all.state.dataset[[current.state.name]], 1)$results),
  row.names = last(all.state.dataset[[current.state.name]])$name)

# second option
# using <- mail.votes

remaining <- data.frame(exp.remaining = pmax(0, by.county$tot_exp_vote - rowSums(data.matrix(all.votes))), 
                        row.names = by.county$name)
ci.mail <- merge(remaining, CI.by.county(using, 0.95), by=0)
ci.other <- merge(remaining, CI.by.county(all.votes, 0.95), by=0)
ci.other <- ci.other[ci.other$Row.names %in% 
                       setdiff(ci.other$Row.names, ci.mail$Row.names), ]
est <- rbind(ci.mail, ci.other)
rm(remaining)

est <- est[order(est$Row.names), ]
rownames(est) <- est$Row.names
est$Row.names <- NULL
est[est$exp.remaining > 0, ]
```


## Estimate the final range

With the lower and upper bound of Biden's and Trump's probability in a county,
we use both probability to calculate the CI. 
The lower end of the CI from the low probability is a candidate's lower bound,
while the upper end of the CI from the high probility is a candidate's upper bound.
We then see the margin using Biden's lower bound $-$ Trump's upper bound, 
and using Biden's upper bound $-$ Trump's lower bound, to calculate the final projection interval.

```{r}
future.lo <- range_est(est$exp.remaining, est$bidenj.low)[1] - range_est(est$exp.remaining, est$trumpd.high)[2]
future.hi <- range_est(est$exp.remaining, est$bidenj.high)[2] - range_est(est$exp.remaining, est$trumpd.low)[1]
current.diff <- sum(all.votes$bidenj) - sum(all.votes$trumpd)
current.diff + c(future.lo, future.hi)
```



